### 2021-07-07 src/paper/report.tex
Smallish updates.

Since early 2020,\comment[id=J]{Put stronger focus on time dependent rapid test
sensitivity (30\% before day of onset of infectiousness)} the CoViD-19 pandemic has
presented an enormous challenge to humanity on many dimensions. The development of
highly effective vaccines holds the promise of containment in the medium term. However,
most countries find themselves many months---and often years---away from reaching
vaccination-induced herd immunity \citep{Swaminathan2021}.\comment[id=HM]{Need to find
some more scientific citation before submission} In the meantime, it is of utmost
importance to employ an effective mix of strategies for containing the virus. The most
frequent initial response was a set of non-pharmaceutical interventions (NPIs) to reduce
contacts between individuals. While this has allowed some countries to sustain
equilibria with very low infection numbers,\footnote{See \citet{Contreras2021} for a
theoretical equilibrium at low case numbers which is sustained with
test-trace-and-isolate policies.} most have seen large fluctuations of infection rates
over time. Containment measures have become increasingly diverse and now include rapid
testing, more nuanced NPIs, and contact tracing. Neither these policies' effect nor the
influence of seasonal patterns or of more infectious virus strains are well understood
in quantitative terms.

@@

The main issue with official case numbers is that they will contain only a fraction of
all infections. We thus model official cases as depicted in
Figure~\ref{fig:model_official_cases}. We take aggregate estimates of the share of
detected cases and use data on \replaced[id=K]{the share of PCR tests administered to
people with CoViD-19 symptoms}{whether CoViD-19 symptoms led to a PCR test}. As the share
of asymptomatic individuals varies by age group, this gives us age-specific shares (see
Figure~\ref{fig:share_known_cases_by_age_group} for the share of known cases by age group
over time in our model). Our estimates suggest that---in the absence of rapid
testing---the detection rate is 80\% higher on average for individuals above age 80
compared to school age children. Once rapid test become available, confirmation of a
positive result is another reason leading to positive PCR tests.

The model is applied to the second and third wave of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021.
Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_infectious_contacts} the
Oxford Response Stringency Index \citep{Hale2020}, which tracks the tightness of
non-pharmaceutical interventions. The index is shown for illustration of the NPIs, we
never use it directly. For legibility reasons, we transform the index so that lower
values represent higher levels of restrictions. A value of zero means all measures
incorporated in the index are turned on. The value 1 represents the situation in
mid-September, with restrictions on gatherings and public events, masking requirements,
but open schools and workplaces. In the seven weeks between mid September and early
November, cases increased by a factor of 10. Restrictions were somewhat tightened in
mid-October and again in early November. New infections remained constant throughout
November before rising again in December, prompting the most stringent lockdown to this
date. Schools and daycare centers were closed again, so were customer-facing businesses
except for grocery and drug stores. From the peak of the second wave just before
Christmas until the trough in mid-February, newly detected cases decreased by almost
three quarters. The third wave in the spring of 2021 is associated with the B.1.1.7
strain, which became dominant in March (Figure~\ref{fig:share_b117}). In early March,
some NPIs were being relaxed; e.g., hairdressers and home improvement stores were
allowed to open again to the public. There were many changes in details of regulations
afterwards, but they did not change the stringency index.

@@

By this time, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}). By the end of May, just
over 40\% had received at least one dose of a vaccine. Around the same time, rapid tests
started to replace regular PCR tests for staff in many medical and nursing facilities.
These had to be administered by medical doctors or in pharmacies. At-home tests approved
by authorities became available in mid-March, rapid test centers were opened, and one
test per person and week was made available free of charge. In several states, customers
were only allowed to enter certain stores with a recent negative rapid test result.
These developments are characteristic of many countries: The initial focus on NPIs to
slow the spread of the disease has been accompanied by vaccines and a growing acceptance
and use of rapid tests. At broadly similar points in time, novel strains of the virus
have started to pose additional challenges.

@@

The effects of various mechanisms can be disentangled due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_infectious_contacts} summarize how contact reductions,
increased hygiene regulations, and seasonality evolved since early September for each of
the three broad contact networks. For example, a value of 0.75 for the work multiplier
means that if the environment was the same as in September (levels of infection rates, no
rapid tests or vaccinations, only the wildtype virus present), infections at the
workplace would be reduced by 25\%. The lines show the product of the effect of contact
reductions, increased hygiene regulations, and seasonality. Two aspects are particularly
interesting. First, all lines broadly follow the stringency index and they would do so
even more if we left out seasonality and school vacations (roughly the last two weeks of
October, two weeks each around Christmas and Easter, and some days in late May). Second,
the most stringent regulations are associated with the period of strong decreases in new
infections between late December 2020 and mid-February 2021. The reversal of the trend is
associated with the spread of the B.1.1.7 variant. The steep drop in recorded cases
during May 2021 is associated with at least weekly rapid tests for 42\% of the
population, a vaccination rate that rose from 28\% to 43\%, and \deleted[id=K]{mostly}
seasonality \replaced[id=K]{further lowering}{impacting a fall in} the relative
infectiousness of contacts \deleted[id=K]{outside of work and school.}

@@

Until mid-March, there is no visible difference between the different scenarios.
Seasonality hardly changes, and only few vaccinations and rapid tests were administered.
Even thereafter, the effect of the vaccination campaign is surprisingly small at first
sight. Whether considering recorded or total infections with only one channel active,
the final level is always the highest in case of the vaccination campaign (orange
lines). The Shapley value decomposition shows that vaccinations contribute 16\% to the
cumulative difference between scenarios. Reasons for this are the slow start---it took
until March~24th until 10\% of the population had received their first vaccination, the
20\% mark was reached on April 19th---and the focus on older individuals. These groups
contribute less to the spread of the disease than others due to a lower number of
contacts\replaced[id=K]{as they do not work, do not go to school and tend to live in small households}{, see Supplementary Material~\ref{subsec:contacts_by_age}}.
% We don't draw age specific number of contacts because the sample sizes would be to
% small.
By the end of our study period, when first-dose vaccination rates reached around 40\% of
the population, the numbers of new cases would have started to decline. It is important
to note that the initial focus of the campaign was to prevent deaths and severe disease.
Indeed, the case fatality was rate considerably lower during the third wave when compared
to the second (4.4\% between October and February and 1.4\% between March and the end of
May).\comment[id=K]{This needs a source.}

>>>>

Since early 2020, the CoViD-19 pandemic has presented an enormous challenge to humanity
on many dimensions. The development of highly effective vaccines holds the promise of
containment in the medium term. However, most countries find themselves many
months---and often years---away from reaching vaccination levels that would end the
pandemic or even protect the most vulnerable\citep{Swaminathan2021}.\comment[id=HM]{Use
\url{https://www.nature.com/articles/s41562-021-01160-2} instead} In the meantime, it is
of utmost importance to employ an effective mix of strategies for containing the virus.
The most frequent initial response was a set of non-pharmaceutical interventions (NPIs)
to reduce contacts between individuals. While this has allowed some countries to sustain
equilibria with very low infection numbers,\footnote{See \citet{Contreras2021} for a
theoretical equilibrium at low case numbers which is sustained with
test-trace-and-isolate policies.} most have seen large fluctuations of infection rates
over time. Containment measures have become increasingly diverse and now include rapid
testing, more nuanced NPIs, and contact tracing. Neither these policies' effect nor the
influence of seasonal patterns or of more infectious virus strains are well understood
in quantitative terms.

@@

The main issue with official case numbers is that they will contain only a fraction of
all infections. We thus model official cases as depicted in
Figure~\ref{fig:model_official_cases}. We take aggregate estimates of the share of
detected cases and use data on the share of PCR tests administered because of CoViD-19
symptoms. As the share of asymptomatic individuals varies by age group, this gives us
age-specific shares (see Figure~\ref{fig:share_known_cases_by_age_group} for the share
of known cases by age group over time in our model). Our estimates suggest that---in the
absence of rapid testing---the detection rate is 80\% higher on average for individuals
above age 80 compared to school age children. Once rapid test become available,
confirmation of a positive result is another reason leading to positive PCR tests.

The model is applied to the second and third waves of the CoViD-19 pandemic in Germany,
covering the period mid-September 2020 to the end of May 2021.
Figure~\ref{fig:pandemic_drivers_model_fit} describes the evolution of the pandemic and
of its drivers. The black line in Figure~\ref{fig:aggregated_fit} shows officially
recorded cases; the black line in Figure~\ref{fig:stringency_infectious_contacts} the
Oxford Response Stringency Index \citep{Hale2020}, which tracks the tightness of
non-pharmaceutical interventions. The index is shown for illustration of the NPIs, we
never use it directly. For legibility reasons, we transform the index so that lower
values represent higher levels of restrictions. A value of zero means all measures
incorporated in the index are turned on. The value 1 represents the situation in
mid-September, with restrictions on gatherings and public events, masking requirements,
but open schools and workplaces. In the seven weeks between mid September and early
November, cases increased by a factor of 10. Restrictions were somewhat tightened in
mid-October and again in early November. New infections remained constant throughout
November before rising again in December, prompting the most stringent lockdown to this
date. Schools and daycare centers were closed, so were customer-facing businesses except
for grocery and drug stores. From the peak of the second wave just before Christmas
until the trough in mid-February, newly detected cases decreased by almost three
quarters. The third wave in the spring of 2021 is associated with the B.1.1.7 strain,
which became dominant in March (Figure~\ref{fig:share_b117}). In early March, some NPIs
were being relaxed; e.g., hairdressers and home improvement stores were allowed to open
again to the public. There were many changes in details of regulations afterwards, but
they did not change the overall stringency index.

@@

By March 2021, the set of policy instruments had become much more diverse. Around the
turn of the year, the first people were vaccinated with a focus on older age groups and
medical personnel (Figure~\ref{fig:antigen_tests_vaccinations}). Until the end of May,
43\% had received at least one dose of a vaccine. In late 2020, rapid tests started to
replace regular PCR tests for staff in many medical and nursing facilities. These had to
be administered by medical doctors or in pharmacies. At-home tests approved by
authorities became available in mid-March, rapid test centers were opened, and one test
per person and week was made available free of charge. In several states, customers were
only allowed to enter certain stores with a recent negative rapid test result. These
developments are characteristic of many countries: The initial focus on NPIs to slow the
spread of the disease has been accompanied by vaccines and a growing acceptance and use
of rapid tests. At broadly similar points in time, novel strains of the virus have
started to pose additional challenges.

@@

The effects of various mechanisms can be disentangled due to the distinct temporal
variation in the drivers of the pandemic. Next to the stringency index, the three lines
in Figure~\ref{fig:stringency_infectious_contacts} summarize how contact reductions,
increased hygiene regulations, and seasonality evolved since early September for each of
the three broad contact networks. For example, a value of 0.75 for the work multiplier
means that if the environment was the same as in September (levels of infection rates,
no rapid tests or vaccinations, only the wildtype virus present), infections at the
workplace would be reduced by 25\%. The lines show the product of the effect of contact
reductions, increased hygiene regulations, and seasonality. Two aspects are particularly
interesting. First, all lines broadly follow the stringency index and they would do so
even more if we left out seasonality and school vacations (roughly the last two weeks of
October, two weeks each around Christmas and Easter, and some days in late May). Second,
the most stringent regulations are associated with the period of strong decreases in new
infections between late December 2020 and mid-February 2021. The reversal of the trend
is associated with the spread of the B.1.1.7 variant. The steep drop in recorded cases
during May 2021 is associated with at least weekly rapid tests for 42\% of the
population, a vaccination rate that rose from 28\% to 43\%, and seasonality lowering the
relative infectiousness of contacts, most strongly outside of work and school.

@@

Until mid-March, there is no visible difference between the different scenarios.
Seasonality hardly changes, and only few vaccinations and rapid tests were administered.
Even thereafter, the effect of the vaccination campaign is surprisingly small at first
sight. Whether considering recorded or total infections with only one channel active,
the final level is always the highest in case of the vaccination campaign (orange
lines). The Shapley value decomposition shows that vaccinations contribute 16\% to the
cumulative difference between scenarios. Reasons for this are the slow start---it took
until March~24th until 10\% of the population had received their first vaccination, the
20\% mark was reached on April 19th---and the focus on older individuals. These groups
contribute less to the spread of the disease than others due to a lower number of
contacts see Supplementary Material~\ref{subsec:contacts_by_age}. By the end of our
study period, when first-dose vaccination rates reached around 40\% of the population,
the numbers of new cases would have started to decline. It is important to note that the
initial focus of the campaign was to prevent deaths and severe disease. Indeed, the case
fatality was rate considerably lower during the third wave when compared to the second
(4.4\% between October and February and 1.4\% between March and the end of
May).\comment[id=K]{This needs a source.}
